---
title: "COURSERA PROJECT 1"
author: "Elshan Musazade"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
# Loading necessary libraries
library(dplyr)
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE)
```

## 🔹 Introduction

This assignment uses data from a personal activity monitoring device that collects steps at 5-minute intervals. The dataset spans October and November 2012 and will be analyzed to explore physical activity patterns, handle missing data, and compare activity behavior between weekdays and weekends.

## 🔹 Load and Inspect Data

```{r load_data}
# Load dataset
activity <- read.csv("C:/Users/ASUS/Desktop/homework/activity.csv")

# Convert date to Date type and check structure
activity$date <- as.Date(activity$date)
str(activity)

# Check for missing data in 'steps'
missing_steps <- sum(is.na(activity$steps))
missing_steps
```

## 🔹 Total Number of Steps per Day

```{r total_steps_day}
# Calculate total steps per day
daily_steps <- activity %>%
  group_by(date) %>%
  summarise(total_steps = sum(steps, na.rm = TRUE))

# Plot histogram of total steps per day
ggplot(daily_steps, aes(x = total_steps)) +
  geom_histogram(binwidth = 1000, fill = "steelblue", color = "black") +
  labs(title = "Histogram of Total Steps per Day", x = "Steps", y = "Frequency") +
  theme_minimal()

# Calculate and display mean and median of daily steps
mean_steps <- mean(daily_steps$total_steps)
median_steps <- median(daily_steps$total_steps)
mean_steps
median_steps
```

## 🔹 Average Daily Activity Pattern

```{r daily_pattern}
# Calculate average steps by 5-minute interval
interval_avg <- activity %>%
  group_by(interval) %>%
  summarise(avg_steps = mean(steps, na.rm = TRUE))

# Plot average steps by interval
ggplot(interval_avg, aes(x = interval, y = avg_steps)) +
  geom_line(color = "darkgreen") +
  labs(title = "Average Steps by 5-Minute Interval", x = "Interval", y = "Average Steps") +
  theme_minimal()

# Find the interval with the maximum average steps
max_interval <- interval_avg[which.max(interval_avg$avg_steps), ]
max_interval
```

## 🔹 Imputing Missing Values

```{r impute}
# Impute missing steps using the average steps for the same interval
activity_imputed <- activity %>%
  left_join(interval_avg, by = "interval") %>%
  mutate(steps = ifelse(is.na(steps), avg_steps, steps)) %>%
  select(date, interval, steps)

# Check for any remaining missing values
remaining_missing <- sum(is.na(activity_imputed$steps))
remaining_missing
```

## 🔹 Total Steps After Imputation

```{r hist_imputed}
# Calculate total steps per day after imputation
daily_steps_imputed <- activity_imputed %>%
  group_by(date) %>%
  summarise(total_steps = sum(steps))

# Plot histogram of total steps after imputation
ggplot(daily_steps_imputed, aes(x = total_steps)) +
  geom_histogram(binwidth = 1000, fill = "coral", color = "black") +
  labs(title = "Histogram After Imputation", x = "Steps", y = "Frequency") +
  theme_minimal()

# Calculate and display mean and median of total steps after imputation
mean_imputed_steps <- mean(daily_steps_imputed$total_steps)
median_imputed_steps <- median(daily_steps_imputed$total_steps)
mean_imputed_steps
median_imputed_steps
```

## 🔹 Compare Before vs After Imputation

```{r compare_means}
# Compare mean and median before and after imputation
original_stats <- daily_steps %>%
  summarise(mean = mean(total_steps), median = median(total_steps))

imputed_stats <- daily_steps_imputed %>%
  summarise(mean = mean(total_steps), median = median(total_steps))

# Display comparison
comparison <- rbind(original_stats, imputed_stats)
comparison
```

## 🔹 Weekday vs Weekend Patterns

```{r weekday_weekend}
# Add 'day_type' column to classify days as 'weekday' or 'weekend'
activity_imputed$day_type <- ifelse(weekdays(activity_imputed$date) %in% c("Saturday", "Sunday"), "weekend", "weekday")

# Calculate average steps by interval for weekdays and weekends
interval_daytype <- activity_imputed %>%
  group_by(interval, day_type) %>%
  summarise(avg_steps = mean(steps), .groups = 'drop')

# Plot steps by interval for weekdays vs weekends
ggplot(interval_daytype, aes(x = interval, y = avg_steps, color = day_type)) +
  geom_line() +
  facet_wrap(~day_type, nrow = 2) +
  labs(title = "Steps by Interval: Weekday vs Weekend", x = "Interval", y = "Average Steps") +
  theme_minimal()
```

## ✅ Conclusion

This report explored activity patterns, handled missing values with mean imputation, and compared behaviors between weekdays and weekends. The analysis was conducted using modern `dplyr` and `ggplot2` functions for efficient and reproducible results.
